\documentclass[]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
%% \usepackage{listings}
%% auto break lines
%% \lstset{breaklines=true}

\usepackage{tikz} % necessary for rendering the binary tree visualization
\usetikzlibrary{positioning}
\usetikzlibrary{shapes, arrows}
\usepackage{hyperref} % better referencing
\usepackage{xcolor} % for better highlighting in tables
\usepackage{longtable} % for appendix
\usepackage{csvsimple} % for appendix

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2026}
\copyrightclause{Copyright for this paper by its authors.
  Use permitted under Creative Commons License Attribution 4.0
  International (CC BY 4.0).}

%%
%% This command is for the conference information
\conference{10th Workshop on Practical Aspects of Automated Reasoning,
Lisboa, 2026}

%%
%% The "title" command
\title{Efficient Multi-Scale Indexing with Dynamic IntMaps}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.

\author[1]{Albert Eisfeld}[%
email=inf23040@lehre.dhbw-stuttgart.de,
]

\author[1]{Stephan Schulz}[%
orcid=0000-0001-6262-8555,
email=schulz@eprover.org,
]

\address[1]{DHBW Stuttgart, Stuttgart, Germany}
%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  Term and clause indexing are central technologies for efficient
  first-order theorem provers and similar reasoning systems. While
  there are a selection of basic technologies, most of them require an
  efficient mapping from function symbols (usually encoded as small
  integers) or other integer values to branches of a tree
  structure. With today's wide variety of application problems, both
  the range of possible values and the cardinality of each individual
  map vary wildly. We present a data structure that is designed to be
  efficient in time and space at all scales and key
  distributions. Experiments show very good scalability and a modest
  improvement over the old, already quite refined data structure in E.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
  term indexing \sep
  data structures \sep
  theorem proving \sep
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

Most powerful modern theorem provers like E~\cite{SCV:CADE-2019},
Vampire~\cite{BBCHHHKRRRSSV:CAV-2025} and iProver~\cite{DK:IJCAR-2020}
are based on saturating calculi, in particularly variations of
superposition~\cite{BG94} and related calculi. These calculi represent
the proof state as a set of clauses. They enrich the proof state by
adding logical consequences of existing clauses, with the aim of
eventually generating the empty clause as an explicit witness of
unsatisfiability (and hence conclude a proof by contradiction). In
addition to generating inferences, simplifying inferences like
subsumption and rewriting are critical for success. These techniques
either remove redundant clause, or replace them by simpler ones, and
typically take up most of the time spent by the inference engine.

Successful provers combine efficient execution and intelligent proof
search heuristics to achieve their power. Especially for harder
problems with large search states, \emph{term indexing} is a critical
technology. Indexing of terms (and of clauses) allows a system to
quickly identify potential inference partners for generating
inferences like paramodulation/superposition or resolution, but in
particular also for simplification techniques like rewriting and
subsumption.

Most modern term indices are based on (often trie-like) trees. At each
node in the tree, branches labelled with function symbols (or
variables) lead to sub-trees. A complete branch of the tree represents
a query compatible with a certain query, and the leaves represent sets
of terms (or clauses) that are potential answers to the this query. In
this paper, we are in particular concerned with the efficiency of
mapping function (or variable) symbols to tree branches. However, the
developed solutions have more general applications, both in indexing
and beyond.

\section{Background and Motivation}

Early contributions to indexing were made by
Stickel~\cite{Stickel:TRIndexing-89}, who introduced \emph{path
  indexing}, and by McCune, who integrated path indexing and
\emph{discrimination tree indexing} into Otter~\cite{Mc92}, creating
the first successful modern saturating prover. Our own prover E uses
\emph{perfect discrimination tree indexing} for forward rewriting,
\emph{fingerprint indexing}~\cite{Schulz:IJCAR-2012} for backwards
rewriting and superposition/paramodulation, and \emph{feature vector
  indexing}~\cite{Schulz:FVI-2013} for subsumption and related
techniques.

When indexing techniques were first developed, most problems for
automated theorem provers were handcrafted, and small in number of
axioms and signature size. Indexing data structures, mapping (mostly)
from function symbols to branches of the index tree, hence usually had
a low branching factor and also a very limited range of values to
branch on. Implementations of these maps used either (potentially
dynamic) arrays with direct mapping, or linear lists with linear
search. Arrays are faster, but if the node is not densely populated,
may waste space. Linear lists have a higher per-key cost, both in
speed and memory, but handle sparse key spaces efficiently.

In E, our original implementation used two dynamic arrays to handle
each node - one for variable-indexed branches, one for function-symbol
indexed branches. This worked well for most problems of the
time. However, in 2003, Urban was converting the MIZAR
library~\cite{TB:IJCAI-1985} to first-order
logic~\cite{Urban:MKM-2003,Urban:JAR-2004} and began experiments
running E on the resulting problems. MIZAR problems have many
thousands of function symbols, and the resulting large, but sparsely
populated arrays led to an explosion of memory consumption by E's
perfect discrimination tree indices (at that time the only indexing
data structure used). Urban solved this by hacking the prover and
replacing the function symbol array with a binary search tree. While
this solved his immediate problem, it decreased performance for other
classes of problems, and was the optimal solution for all situations.

However, inspired by Urban's work, we replaced both the original
solution and Urban's hack with a dedicated \texttt{IntMap} data
type. It's abstract interface supports inserting and deleting
key/value pairs, querying the value for a given key, and iteration over
all or an arbitrary sub-range of key/value pairs stored in a given map.

Before we discuss the details of the original and the new
\texttt{IntMap}, we present a couple of observations that shape the
design.

First, the proof state of saturating theorem provers tends to increase
in size over time, and so does the number of indexed clauses. There
are occasional cases where already indexed clauses become redundant
and have to be removed from the index (\emph{backward
  simplification}), but this is a rare occurrence. Typically, all
indices (and all \texttt{IntMaps} used in them) tend to grow over
time.

The most frequent operation by far is the query for a given key (and
its associated value). Insertions are orders of magnitude more
frequent than deletions. In-order iteration over key/value pairs is a
frequent operation as well - more frequent than insertions, but less
frequent than look-ups.

Our normal use case are term and clause indices. These are typically
arranged as trees, with each node of the tree representing a set of
terms or clauses, and each branching splitting this set into disjoint
subsets. Each branch is typically labelled by a small integer (the
\emph{key}), often representing the encoding of a function symbol
(i.e.\ its index in the signature table) or a variable (typically a
small negative number), and resulting from test on the terms in that
branch. As a result, the sets (and the branching factors) of the early
nodes are often quite large, but as we follow the branches, the
subsets and hence the branching factors get smaller and
smaller. Earlier nodes are accessed much more often (dominating access
times), but there are many more nodes deeper in the tree, which are
thus contributing significantly to memory use.

While the branching factor goes down with depth, the range (i.e.\ the
difference between the smallest and the largest key) only goes down
statistically, and not to the same degree. We have fewer keys, but
drawn from the same total range. This range varies a lot between
problems. Variables are typically restricted to the low
dozens. Function symbols (and predicate symbols) can vary from single
operators to tens of thousands for problems from Mizar and
SUMO~\cite{NP:FOISS-2001}, and over a million for example problems
from Cyc~\cite{RRG:CO-2005}.

These observations lead us to the conclusion that the mapping of keys
(integers) to branches in the indices must be efficient in time and
memory at many different scales.

\section{The Evolution of IntMap}

As stated above, our first implementation of branching in perfect
discrimination trees relied on two separate dynamic arrays (a generic
datatype already implemented in E's libraries), with direct mapping
from positive \texttt{long int} keys to \texttt{void*} values. These
supported only 0-based indexing. Urban replaced the function symbol
side of the index by \texttt{NumTree}s, another of E's built-in data
types. \texttt{NumTree}s are one of several implementations of Splay
trees~\cite{ST:JACM-85} in E, in particular Splay trees that use
\texttt{long int} keys and either \texttt{long int} or \texttt{void*}
values. Splay trees are structurally normal binary search trees, and
can be used as such. However, most operations that require key lookup
rearrange the tree, to move the requensted key or one of its
neighbours to the root of the tree. As a result, Splay trees provide
an amortised logarithmic cost guarantee, and in particular, if only a
subset of $n$ keys is frequently accessed, essentially behave like a
tree of only these $n$ keys.

We then introduced \texttt{IntMap} as an abstract data type. It
supports insertion, deletion, lookup, and ordered iteration via an
iterator data type with \emph{allocate}, \emph{next} and \emph{free}
operations.

This \texttt{IntMap} datatype had a polymorphic implementation, with 4
distinct cases:
\begin{itemize}
\item \emph{Empty}: The initial state on creation. No key/value pair
  is stored. The iterator for an empty \texttt{IntMap} simply returns
  an unspecified key and value \texttt{NULL}, the marker for the end
  of iteration.
\item \emph{Single}: A frequent case in indexing are long branches
  created by a single indexed term. To accommodate this case
  efficiently, we provide an implementation that stores a single key
  and associated value. The iterator for this case uses a boolean value
  to indicate if this one key/value pair has been seen already. If
  not, it is returned, if so, it returns \texttt{NULL} (with an
  unspecified key) again.
\item \emph{Array}: Dense populations of multiple keys are stored in a
  \texttt{PDRangeArray}. This is a flexible dynamic array type from
  E's library. It allows indexing with \texttt{long int} values
  starting at an arbitrary (potentially negative) offset. This array
  automatically grows as needed, but never shrinks (remember that
  deletions are rare events). The iterator for the array case
  maintains the index of the next unseen array element. The
  \emph{next} operation searches for the next non-\texttt{NULL} value
  in the array and returns the corresponding key/value pair, or
  \texttt{NULL} if runs out of array elements.
\item \emph{Tree}: Sparse key/value populations are stored in a
  \texttt{NumTree}, as described above. The iterator uses a stack to
  describe the current state of iteration, and uses a non-destructive
  in-order (Left-Node-Right) traversal to access all nodes. It returns the stored
  key/value pair at each node, or \texttt{NULL} if it runs out of
  nodes.
\end{itemize}

In both of the multi-key-modes, the \texttt{IntMap} monitors
population density. If the density falls below a certain threshold
value, and the map is in array mode, it reorganises itself into tree
mode. If the density in tree mode increases beyond a second (larger)
threshold, it switches from tree to array mode. The difference in the
two thresholds provides a hysteresis that avoids overly frequent
reorganisations. In our implementation, we switch to array if more
than one in 8 spaces are occupied, and we switch to tree if less than
one in 4 spaces are occupied. When first switching to multi-key-form
with density in the hysteresis range, array is preferred (on the
assumption that the density is more likely to increase than to fall).

\subsection{Recent Improvements}

Observations have shown that the key-value-pairs often lie in the
memory as clusters, so that there are some smaller regions which are
significantly denser populated than the overall set of all
key-value-pairs considered for a tree. A possible reason for this is
that in larger specifications, there are sub-theories mostly dealing
with a subset of functions and relations, while also being tied into
the common theory with some shared symbols. This is then reflected in
indexing data structures for clauses from these sub-theories.

To accomodate this circumstance, the structure of the \texttt{IntMap}
datatype was simplified to use only one case instead of the two
\emph{Array} and \emph{Tree} for multi-key-modes, combining both of
the older cases into one. Key/Value pairs are stored in a tree of
smaller arrays. This new \emph{Array-Tree} is based on the
\texttt{NumXTree} datatype, a binary tree that contains arrays with a
size of $2^n$ entries in each node. Keys are effectively split into
two parts, one part (the high bits) selecting the array in the tree,
the other (the low $n$ bits) the index of the value in the array.

\autoref{fig:NumXTree} shows the visualisation of an exemplary set of
key-value-pairs.  The number next to the node represents the key of
the tree node and thus the high bits of each key in that array. The
index in the array represents the low 2 bits of the key.

\begin{figure}[h!]
  \centering
  \begin{tikzpicture}
    [
      treenode/.style = {shape=rectangle, draw, align=center, rounded corners=3pt},
      level 1/.style = {sibling distance=8cm},
      level 2/.style = {sibling distance=4cm},
      level 3/.style = {sibling distance=2cm},
      index/.style = {color=red, font=\footnotesize\bfseries, inner sep=1pt}
    ]
    \node (root) [treenode] {\texttt{[-, a, b, -]}}
    child { node (L) [treenode] {\texttt{[j, -, -, -]}}
        child { node (LL) [treenode] {\texttt{[-, -, c, -]}} }
        child { node (LR) [treenode] {\texttt{[d, e, -, g]}} }
    }
    child { node (R) [treenode] {\texttt{[-, k, l, m]}}
        child { node (RL) [treenode] {\texttt{[f, -, n, -]}} }
        child { node (RR) [treenode] {\texttt{[-, h, i, -]}} }
    };

    \node [index, left=2pt of root.south west] {16};
    \node [index, left=2pt of L.south west]    {4};
    \node [index, left=2pt of LL.south west]   {-4};
    \node [index, left=2pt of LR.south west]   {12};
    \node [index, left=2pt of R.south west]    {24};
    \node [index, left=2pt of RL.south west]   {20};
    \node [index, left=2pt of RR.south west]   {40};

  \end{tikzpicture}
  \caption{
    Visualisation of exemplary \texttt{NumXTree} binary tree with an
    array size of 4. Values in the array that are marked with
    '\texttt{-}' represent \texttt{NULL}-values, i.e.\ these keys
    are not associated with any value.
    }
  \label{fig:NumXTree}
\end{figure}

Using the \texttt{NumXTree} binary tree, the new \texttt{IntMap}
datatype still supports insertion, deletion, lookup and ordered
iteration as in the older version.

Similar to the \emph{Tree} case, the ordered iteration uses a
non-destructive in-order traversal. A stack is used to keep track of
the current array in the tree. Because of the nested structure,
iterating over the tree nodes alone would yield only values at
index 0 of the array. To iterate over all key-value-pairs, the
iterator uses an additional index counter to indicate the next unused
entry in the current array. Iteration is hence done at two levels:
First, the iterator looks for the next occupied entry in the current
array. If it runs out of possibilities, the iterator moves to the next
tree node and resets the internal index counter. In terms of memory,
the iterator combines both the stack and the array index of the old
model, thus using slightly more memory. However, iterators are are, % <- likely an error :)
and the (dynamic) stack can typically be smaller, since there are
fewer tree nodes.

Using the exemplary tree from \autoref{fig:NumXTree}, an iterator
would return the values in the following order: \texttt{$\uparrow$c,
  $\uparrow$j, d, e, g, $\uparrow$a, b, $\uparrow$f, n, $\uparrow$k,
  l, m, $\uparrow$h, i}, where each of the $\uparrow$-occurances marks
transition to a new tree node. The values have the keys depicted
in \autoref{tab:key_value_pairs}.


\begin{table}[h!]
  \centering
    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|l|l|l|l|l|}
      \hline
      \textbf{Value} & c & j & d & e & g & a & b & f & n & k & l & m & h & i \\ \hline
      \textbf{Key} & -2 & 4 & 12 & 13 & 15 & 17 & 18 & 20 & 22 & 25 & 26 & 27 & 41 & 42  \\ \hline
    \end{tabular}
  \caption{
    Key-value-pairs of exemplary \texttt{NumXTree} binary tree
    in \autoref{fig:NumXTree}
    }
  \label{tab:key_value_pairs}
\end{table}

The proposed \texttt{NumXTree} binary tree comes with benefits for
memory and runtime. Each node now can store multiple key-value-pairs
while still needing only two words of memory for the left and right
child-nodes to maintain the tree structure.  Furthermore, the amount
of operations for rearranging the tree is reduced on average, since
fewer nodes can hold the same number of values. Additionally,
reorganising the \texttt{IntMap} when the critical density is reached
is now obsolete, since there is only one multi-key-mode.

Since the structure is optimised on the common case, it is expected to
perform worse than the original structure in runtime and memory in
edge cases for the key/value distribution. If the key-value-pairs are
either very densely or very sparsely populated, organising the pairs
in an array or a binary tree respectively will be optimal for memory
and in the case of the array state, also in runtime. However, on
average we expect better behaviour, and in particular better
worst-case behaviour.

\section{Experimental Results}

% No big speed gain, but typically lower memory consumption, and no
% negative outliers?
To examine whether the changed data structure actually improves prover
runtime and memory footprint, we ran the prover on various problems
from the TPTP library~\cite{Sutcliffe:JAR-2017}, version 9.2.1. In
order to provide a sufficient comparison, the theorem prover was built
using the original state of the \texttt{IntMap} data structure with
two states for organising multiple key-value-pairs, as well as six
versions of the improved \texttt{IntMap}, with array-sizes of 1, 2, 4,
8, 16 and 32.

All versions of the program were built using the gcc compiler with the
optimisation flags \linebreak \texttt{-O03 -fno-common}.  The tests
have been done on a Linux server with Intel Xeon Platinum 8176 CPUs
running at 2.10GHz under Ubuntu 24.04.3 LTS. The machine had 1.5TB
pysical RAM. The source code is available \href{https://github.com/aallbert/eprover}{here}.

To obtain the time used to find the proof for the according problem,
the Linux \texttt{time} command was used. All problems have been run
on their own core with a soft runtime limit of 300 seconds to ensure
consistency and controlled termination.  The command flags used are
the following: \texttt{time taskset -c 32-63 ./eprover --auto
  --detsort-new --detsort-rw --soft-cpu-limit=300 --cpu-limit=350
  <problem>}

% How did we arrive at the probem set? TPTP competition (which years,
% which categories?) to give us problems of adequate difficulty?
We picked 158 problems from the First-order Theorems category of the CADE ATP System
Competition~\cite{Sutcliffe:AI-2016} (see~\autoref{tab:all_problems})
(years 2025 - 2019) to give us a good selection of problems with a good
distribution of difficulties.

From these problems, we excluded all problems that finished below 5
seconds for all versions, to avoid measurement noise. Problem runs
over 300s indicate that the prover did not find a proof in time and
were also excluded from analysis.

This leaves 34 problems with valid data, where at least one version of
the prover runs in between 5 and 300 seconds. The times for these
problems are shown in \autoref{tab:filtered_results}. The
number of times how often each version was the fastest in finding a
proof and the total runtime for all 34 problems are depicted in
\autoref{fig:results}.

\begin{table}[h!]
  \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \begin{tabular}{l|l}
            \textbf{Array Size} & \textbf{Fastest Version} \\ \hline
            4 & 11 \\ \hline
            2 & 10 \\ \hline
            \textit{Standard Version} & 9 \\ \hline
            16 & 2 \\ \hline
            1 & 1 \\ \hline
            8 & 1 \\ \hline
            32 & 0 \\
        \end{tabular}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \begin{tabular}{l|l}
            \textbf{Array Size} & \textbf{Total Time} \\
                                & \textbf{spent (s)} \\ \hline
            8 & 3158.944 \\ \hline
            16 & 3234.791 \\ \hline
            \textit{Standard Version} & 3314.828 \\ \hline
            1 & 3325.156 \\ \hline
            32 & 3366.841 \\ \hline
            4 & 3390.069 \\ \hline
            2 & 3591.788 \\
        \end{tabular}
    \end{minipage}
  \caption{Number of times each version of the data structure was fastest (left)
    and total time spent on the 34 problems (right). Note that the
    \textit{Standard Version} refers to the \texttt{IntMap} data
    structure that uses the binary tree and array to organise the
    key-value-pairs.
    }
  \label{fig:results}
\end{table}


After looking at the runtime distribution of different functions in
the program using a profiler, a maximum improvement of 2\% was expected,
since all functions related to the \texttt{IntMap} take up roughly 2\% of
the runtime for each problem. \autoref{fig:results} shows that the
differences in total runtime in the right table are unexpectedly
larger than 2\%. Additionally, in the set of 34 valid problems are some
outliers where the time differences in between versions are greater than
80\% as shown in \autoref{tab:filtered_results}. This effect has
occurred before, and indicates that the proof search (which can depend
on in-memory allocation of pointers) took a different part. These
outliers hence are not suitable to evaluate the quality of the data
structures.
%We assume that this is due to different
%allocation of the key-value-pairs across different array sizes which
%can favor or disfavor the order in which the axioms are treated in the
%proving process.

The results show that the versions with an array size of 1 and 32
perform worse than the Standard Version in both metrics. Surprisingly,
the number of times a version solved a problem the fastest does not
correlate to its overall runtime. As demonstrated in
\autoref{fig:results}, the versions with an array size of 2 and 4 score
high in the number of times they are the fastest, but low in the total
time they spent on the problems. Note that versions 2 and 4 have more
negative and positive outliers, i.e. they reach the 300 seconds run
time limit where other versions complete the problem in under 120
seconds and vice versa. \autoref{fig:results_no_outliers} shows that
when excluding positive and negative outliers, versions 2 and 4
perform much better in overall runtime compared to the Standard Version.
An array size of 8 provides a stable version that performs best in
runtime with all valid problems included and outperforms the Standard
Version in runtime even when excluding the outliers.

For the 34 valid problems, the memory consumption, i.e. how much memory
was allocated to find a proof, was also monitored. The results are
shown in \autoref{fig:results_memory}.

\begin{table}[h!]
  \centering
    \begin{minipage}{0.45\textwidth}
        \centering
        \begin{tabular}{l|l}
            \textbf{Array Size} & \textbf{Least Memory} \\ \hline
            \textit{Standard Version} & 20 \\ \hline
            8 & 9 \\ \hline
            32 & 4 \\ \hline
            16 & 1 \\ \hline
            4 & 0 \\ \hline
            2 & 0 \\ \hline
            1 & 0 \\
        \end{tabular}
    \end{minipage}
    \begin{minipage}{0.45\textwidth}
        \centering
        \begin{tabular}{l|l}
            \textbf{Array Size} & \textbf{Total Memory} \\
                                & \textbf{allocated (GB)} \\ \hline
            \textit{Standard Version} & 4198.438 \\ \hline
            8 & 4360.340 \\ \hline
            32 & 4415.674 \\ \hline
            16 & 4468.586 \\ \hline
            4 & 4470.512 \\ \hline
            2 & 4561.600 \\ \hline
            1 & 4585.171 \\
        \end{tabular}
    \end{minipage}
  \caption{
    Number of times each version of the data structure consumed the
    least amount of Memory (left) and total memory allocated on the
    34 valid problems (right). Note that the
    \textit{Standard Version} refers to the \texttt{IntMap} data
    structure that uses the binary tree and array to organise the
    key-value-pairs.
    }
  \label{fig:results_memory}
\end{table}

None of the improved versions show a fewer overall memory consumption
compared to the standard version of \texttt{IntMap}. This suggests that the
timesavement does not directly correlate with the amount of memory
consumed and is more dependend on functions that require heavy computation.

\section{Future Work}

Our current implementation is already performing quite well. There are
several options to further improve the data structure, but none that
is likely to make a significant impact in overall prover
performance. The most attractive idea, if primarily from a simplicity
and aesthetics perspective, would be to use dynamically sized array
segments in the current array tree structure, i.e.\ to merge
neighbouring arrays if the resulting larger array would be full
enough, and likewise to split them if the population factor sinks
below a threshold. This would allow an even more smooth transition
between all-array and all-tree representations.

Another idea would be to analyse the problems on typical patterns and
set the array size according to these patterns. Since some problems
show drastic outliers, changing the array size after the prover read
all axioms could have a significant impact on the runtime of some
cases.

\section{Conclusion}

In older work, Nieuwenhuis, Hillenbrand, Riazanov and Voronkov have
found that, back in 2001, the exact choice of indexing mechanism for a
given retrieval operation was less important than the decision to pick
any over none~\cite{NHRV:IJCAR-2001}, and that details of
implementation are not usually critical. Urban's experience shows that
this did not hold up for all cases if the landscape of problems
addressed by automated theorem provers changes significantly, e.g.\
with the introduction of much larger real-world problems (which lead,
among other things, to the idea of a-priory premise selection and the
\emph{Workshop on Empirically Successful Automated Reasoning in Large
  Theories} (ESARLT)~\cite{SUS:ESARLT-2007}, held with CADE-21 in 2007
in Bremen). However, in this work the idea has reasserted itself - the
new approach shows some advantages, but, at least in our test cases,
they seem fairly marginal in run time.
%%% Check the followwing ;-)
With the improved \texttt{IntMap} data structure, a slight but significant
gain in speed has been achieved when considering all valid test-problems,
though causing an increased memory footprint. Besides its impact on runtime,
the improved \texttt{IntMap} also deliveres a more elegant data structure,
where switching the organisation between a binary tree and an array becomes
obsolete.
% The gains seem to be more significant in terms of memory, but, with
% recent progress in hardware, memory is now much less of a concern for
% theorem provers.
%%%
% Still, we believe developing the new data structure was worth it, if
% only for the improved elegance.

%% The declaration on generative AI comes in effect
%% in Janary 2025. See also
%% https://ceur-ws.org/GenAI/Policy.html
\section*{Declaration on Generative AI}

The author(s) have not employed any Generative AI tools.

%%
%% Define the bibliography file to be used
\bibliography{stsbib.bib}

%%
%% If your work has an appendix, this is the place to put it.

\section{Appendix}

\appendix


\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|l|l|}
    \hline
        \textbf{Problem} & \textbf{Std.} & \textbf{S1} & \textbf{S2} & \textbf{S4} & \textbf{S8} & \textbf{S16} & \textbf{S32} \\\hline
        {\color{blue} ITP015+1.p} & 0.076 & 0.112 & 307.674 & 0.069 & 0.137 & 0.193 & 0.166 \\ \hline
        {\color{blue} ITP013+1.p} & 0.162 & 0.161 & 306.51 & 304.999 & 0.157 & 0.156 & 0.163 \\ \hline
        {\color{blue} NUM461+2.p} & 11.443 & 11.519 & 0.678 & 0.071 & 7.098 & 9.299 & 9.627 \\ \hline
        {\color{blue} SYO606+1.p} & 46.245 & 46.688 & 0.091 & 9.234 & 14.102 & 48.293 & 51.437 \\ \hline
        SWW478+7.p & 5.439 & 5.272 & 5.314 & 5.44 & 5.689 & 5.576 & 6.103 \\ \hline
        SEU348+2.p & 7.152 & 7.235 & 7.217 & 7.237 & 7.433 & 7.478 & 7.935 \\ \hline
        PRO003+2.p & 8.063 & 8.129 & 8.135 & 7.832 & 8.217 & 8.032 & 8.659 \\ \hline
        {\color{blue} GRP746+1.p} & 306.37 & 306.48 & 16.701 & 307.238 & 307.567 & 306.824 & 306.486 \\ \hline
        SWW348+1.p & 18.94 & 19.22 & 19.68 & 18.772 & 19.257 & 19.454 & 20.487 \\ \hline
        SWW350+1.p & 19.193 & 19.398 & 19.694 & 19.161 & 19.657 & 19.557 & 19.944 \\ \hline
        SWW337+1.p & 20.907 & 20.797 & 20.613 & 20.449 & 20.679 & 20.118 & 21.34 \\ \hline
        SWW339+1.p & 21.705 & 22.06 & 22.238 & 21.848 & 21.897 & 22.057 & 22.561 \\ \hline
        GEO283+1.p & 22.931 & 23.052 & 25.116 & 25.043 & 26.072 & 23.628 & 24.997 \\ \hline
        GEO286+1.p & 28.884 & 29.085 & 30.16 & 29.611 & 31.799 & 29.576 & 31.16 \\ \hline
        GRP655+2.p & 33.715 & 34.033 & 31.437 & 33.891 & 34.223 & 33.109 & 34.283 \\ \hline
        GEO281+1.p & 31.721 & 31.889 & 35.234 & 33.036 & 34.847 & 32.933 & 34.539 \\ \hline
        PUZ077+1.p & 50.433 & 50.232 & 35.081 & 42.176 & 45.367 & 48.507 & 52.393 \\ \hline
        GEO168+1.p & 44.031 & 44.452 & 44.123 & 43.539 & 44.053 & 43.94 & 46.782 \\ \hline
        SYO605+1.p & 53.539 & 53.839 & 50.49 & 52.702 & 53.238 & 53.46 & 55.582 \\ \hline
        LCL403+2.p & 57.142 & 58.323 & 58.932 & 57.116 & 58.734 & 57.759 & 59.571 \\ \hline
        {\color{blue} SET671+3.p} & 113.988 & 112.306 & 304.77 & 83.846 & 111.828 & 111.754 & 116.296 \\ \hline
        LCL642+1.010.p & 81.422 & 82.259 & 84.527 & 83.955 & 84.406 & 86.174 & 91.24 \\ \hline
        {\color{blue} SWW470+1.p} & 227.801 & 224.679 & 89.866 & 238.68 & 226.527 & 222.825 & 238.192 \\ \hline
        GRP654+2.p & 95.475 & 96.542 & 89.605 & 90.596 & 92.278 & 95.299 & 99.265 \\ \hline
        GRP664+1.p & 113.49 & 112.5 & 105.474 & 107.302 & 102.901 & 110.015 & 114.988 \\ \hline
        {\color{blue} MGT005+2.p} & 216.325 & 217.314 & 113.361 & 110.072 & 114.986 & 114.329 & 121.136 \\ \hline
        LCL642+1.015.p & 122.743 & 123.628 & 126.232 & 126.597 & 126.383 & 130.866 & 137.397 \\ \hline
        {\color{blue} SWW893+1.p} & 140.615 & 140.882 & 307.357 & 124.601 & 128.324 & 138.03 & 146.766 \\ \hline
        SWW884+1.p & 151.058 & 151.878 & 143.756 & 141.206 & 142.347 & 149.41 & 155.895 \\ \hline
        LCL656+1.015.p & 164.727 & 164.912 & 165.782 & 166.106 & 168.33 & 170.198 & 178.473 \\ \hline
        GEO310+1.p & 222.817 & 224.626 & 216.924 & 235.653 & 242.102 & 229.157 & 239.72 \\ \hline
        SCT117+1.p & 293.299 & 293.225 & 221.799 & 255.493 & 264.382 & 285.535 & 305.842 \\ \hline
        GEO311+1.p & 298.235 & 302.223 & 273.717 & 283.145 & 293.984 & 304.098 & 303.921 \\ \hline
        GEO329+1.p & 284.742 & 286.206 & 303.5 & 303.353 & 299.943 & 297.152 & 303.495 \\ \hline
    \end{tabular}
  \caption{
    Time spent of each version on the 34 valid problems, where at least one
    version is in between 5 and 300 seconds runtime. Problems, where the time
    differences in between the versions are greater than 80\% are seen as
    outliers and marked blue. These outliers have been tested three times each
    and the average time of each triple is counted as the valid data point and presented
    in the table. \textit{Std.} is the Standard Version of the \texttt{IntMap}
    data structure, \textit{S1}, \textit{S2}, etc. are the versions that use
    the \texttt{NumXTree} binary tree with the according array size
    }
  \label{tab:filtered_results}
\end{table}

\begin{table}[t!]
  \centering
    \begin{tabular}{l|l}
      \textbf{Array Size} & \textbf{Total Time spent (s)} \\ \hline
      2 & 2144.780 \\ \hline
      4 & 2211.259 \\ \hline
      8 & 2248.218 \\ \hline
      \textit{Standard Version} & 2251.803 \\ \hline
      1 & 2265.015 \\ \hline
      16 & 2283.088 \\ \hline
      32 & 2376.572 \\
    \end{tabular}
  \caption{
    Total time spent on the valid problems where positive and negative
    outliers are \emph{excluded}.
    }
  \label{fig:results_no_outliers}
\end{table}


\csvreader[
  longtable={|l|l|l|l|l|l|l|l|},
  table head={
    \caption{
      Amount of memory each version allocated in GB for solving each problem - \textit{Std.} is
      the Standard Version of the \texttt{IntMap} data structure,
      \textit{S1}, \textit{S2}, etc. are the versions that use the \texttt{NumXTree}
      binary tree with the according array size
    }\label{tab:filtered_problems_memory}\\\hline
    \textbf{Problem} & \textbf{Std.} & \textbf{S1} & \textbf{S2} & \textbf{S4} & \textbf{S8} & \textbf{S16} & \textbf{S32} \\\hline
    \endfirsthead
    \hline
  },
  late after line=\\\hline
]
{test_results_memory.csv}
{1=\Problem, 2=\MemoryVersionStandard, 3=\MemoryVersionOne, 4=\MemoryVersionTwo, 5=\MemoryVersionFour, 6=\MemoryVersionEight, 7=\MemoryVersionSixteen, 8=\MemoryVersionThirtyTwo}
{\Problem & \MemoryVersionStandard & \MemoryVersionOne & \MemoryVersionTwo & \MemoryVersionFour & \MemoryVersionEight & \MemoryVersionSixteen & \MemoryVersionThirtyTwo}    % }



\csvreader[
  longtable={|l|l|l|l|l|l|l|l|},
  table head={
    \caption{
      Time spent of each version on each problem - \textit{Std.} is
      the Standard Version of the \texttt{IntMap} data structure,
      \textit{S1}, \textit{S2}, etc. are the versions that use the \texttt{NumXTree}
      binary tree with the according array size.
    }\label{tab:all_problems}\\\hline
    \textbf{Problem} & \textbf{Std.} & \textbf{S1} & \textbf{S2} & \textbf{S4} & \textbf{S8} & \textbf{S16} & \textbf{S32} \\\hline
    \endfirsthead
    \hline
  },
  late after line=\\\hline
]
{test_results_time.csv}
{1=\Problem, 2=\TimeVersionStandard, 3=\TimeVersionOne, 4=\TimeVersionTwo, 5=\TimeVersionFour, 6=\TimeVersionEight, 7=\TimeVersionSixteen, 8=\TimeVersionThirtyTwo}
{\Problem & \TimeVersionStandard & \TimeVersionOne & \TimeVersionTwo & \TimeVersionFour & \TimeVersionEight & \TimeVersionSixteen & \TimeVersionThirtyTwo}

\end{document}

%%
%% End of file

%%% Local Variables:
%%% mode: LaTeX
%%% TeX-master: t
%%% End:
